Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-01-08 analysis/paper.tex
Revise species list

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 56 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
Species ranges are predicted to have been on significantly smaller in
the Early Holocene compared to present conditions, with many species
that are found throughout the region's `hilly flanks' today indicated to
have much more restricted distributions centered on the Levant, Cyprus
and Western Anatolia. Ranges shrunk by on average c.~10\% from the
terminal Pleistocene to the Early Holocene, with a further c.~10\%
reduction for the duration of the Younger Dryas. However, modelled
ranges do not reliably predict the observed occurrence of specific
species at archaeological sites. {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Introduction
  \item[$\square$]
    Background

@@

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-model-metrics
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\square$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

\caption{\label{fig-region}Study region (red) and archaeological sites
used to generate modelled flora. Solid circles indicate Neolithic
sites.}

@@

We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
all the identifiable species known to be present at more than three
Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
canonical form specified in the GBIF Backbone Taxonomy
\citep{GBIFSecretariat2023}. So for example occurrences for
\emph{Bolboschoenus maritimus} also include those recorded under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

@@

\begin{verbatim}
Warning: HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
HTML tags found, and they will be removed.
* Set `options(gt.html_tag_check = FALSE)` to disable this check.
\end{verbatim}

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
\setlength{\LTpost}{0mm}

\begin{longtable}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.50\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 15.00pt -2\tabcolsep-1.5\arrayrulewidth}}

\caption{\label{tbl-occ-count}Recorded occurrence of flora considered in
this study in Neolithic and contemporary West Asia}

\toprule
 & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 30.00pt -2\tabcolsep-1.5\arrayrulewidth}}{Occurrences} \\ 
\cmidrule(lr){2-4}
Taxon & Neolithic & (\%) & Present \\ 
\midrule\addlinespace[2.5pt]
\emph{Triticum turgidum dicoccum}

incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 & 44\% & 200 \\ 
\emph{Hordeum spontaneum}

incl. \emph{Hordeum vulgare} & 62 & 43\% & 6912 \\ 
\emph{Triticum monococcum aegilopoides}

incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 & 30\% & 3213 \\ 
\emph{Bolboschoenus maritimus}

incl. \emph{Scirpus maritimus} & 33 & 23\% & 364 \\ 
\emph{Vicia ervilia} & 33 & 23\% & 858 \\ 
\emph{Buglossoides tenuiflora} & 31 & 21\% & 159 \\ 
\emph{Arnebia decumbens} & 25 & 17\% & 240 \\ 
\emph{Buglossoides arvensis} & 25 & 17\% & 251 \\ 
\emph{Medicago radiata} & 21 & 14\% & 883 \\ 
\emph{Androsace maxima} & 20 & 14\% & 130 \\ 
\emph{Vicia orientalis}

incl. \emph{Lens culinaris} & 17 & 12\% & 441 \\ 
\emph{Medicago astroites}

incl. \emph{Trigonella astroites} & 16 & 11\% & 130 \\ 
\emph{Arnebia linearifolia} & 14 & 10\% & 158 \\ 
\emph{Linum bienne}

incl. \emph{Linum usitatissimum} & 11 & 8\% & 274 \\ 
\emph{Gypsophila vaccaria}

incl. \emph{Vaccaria pyramidata} & 11 & 8\% & 326 \\ 
\emph{Carex divisa} & 10 & 7\% & 330 \\ 
\emph{Ficus carica} & 10 & 7\% & 4265 \\ 
\emph{Lathyrus oleraceus}

incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 & 6\% & 2269 \\ 
\emph{Vicia faba} & 8 & 6\% & 2578 \\ 
\emph{Aizoanthemopsis hispanica}

incl. \emph{Aizoon hispanicum} & 7 & 5\% & 303 \\ 
\emph{Bolboschoenus glaucus} & 7 & 5\% & 27\textsuperscript{\textit{*}} \\ 
\emph{Pistacia atlantica} & 7 & 5\% & 928 \\ 
\emph{Polygonum arenarium arenarium}

incl. \emph{Polygonum venantianum} & 7 & 5\% & 5\textsuperscript{\textit{*}} \\ 
\emph{Prosopis farcta} & 7 & 5\% & 1205 \\ 
\emph{Rumex pulcher} & 7 & 5\% & 794 \\ 
\emph{Ammi majus} & 6 & 4\% & 461 \\ 
\emph{Cicer reticulatum}

incl. \emph{Cicer arietinum} & 6 & 4\% & 193 \\ 
\emph{Hordeum bulbosum} & 6 & 4\% & 2036 \\ 
\emph{Polygonum corrigioloides} & 6 & 4\% & 4\textsuperscript{\textit{*}} \\ 
\emph{Salsola kali} & 6 & 4\% & 369 \\ 
\emph{Taeniatherum caput-medusae} & 6 & 4\% & 235 \\ 
\emph{Capparis spinosa} & 5 & 3\% & 4031 \\ 
\emph{Chenopodium album} & 5 & 3\% & 357 \\ 
\emph{Lolium temulentum} & 5 & 3\% & 148 \\ 
\emph{Poa bulbosa} & 5 & 3\% & 1401 \\ 
\emph{Aegilops crassa} & 4 & 3\% & 170 \\ 
\emph{Atriplex prostrata} & 4 & 3\% & 112 \\ 
\emph{Brachypodium distachyon} & 4 & 3\% & 1581 \\ 
\emph{Fumaria densiflora} & 4 & 3\% & 194 \\ 
\emph{Gypsophila pilosa} & 4 & 3\% & 79 \\ 
\emph{Hordeum murinum} & 4 & 3\% & 1342 \\ 
\emph{Lathyrus aphaca} & 4 & 3\% & 2085 \\ 
\emph{Lathyrus sativus} & 4 & 3\% & 297 \\ 
\emph{Secale cereale} & 4 & 3\% & 1153 \\ 
\emph{Triticum durum} & 4 & 3\% & 219 \\ 
\emph{Vitis sylvestris} & 4 & 3\% & 3\textsuperscript{\textit{*}} \\ 
\emph{Adonis flammea} & 3 & 2\% & 79 \\ 
\emph{Avena sterilis} & 3 & 2\% & 50001\textsuperscript{\textit{*}} \\ 
\emph{Bassia arabica} & 3 & 2\% & 230 \\ 
\emph{Bromus sterilis} & 3 & 2\% & 385 \\ 
\emph{Cephalaria syriaca} & 3 & 2\% & 144 \\ 
\emph{Glaucium aleppicum} & 3 & 2\% & 22\textsuperscript{\textit{*}} \\ 
\emph{Helianthemum ledifolium} & 3 & 2\% & 668 \\ 
\emph{Lolium rigidum} & 3 & 2\% & 1464 \\ 
\emph{Phalaris paradoxa} & 3 & 2\% & 370 \\ 
\emph{Quercus ithaburensis} & 3 & 2\% & 1431 \\ 
\emph{Triticum aestivum compactum} & 3 & 2\% & 209 \\ 
\emph{Verbena officinalis} & 3 & 2\% & 1046 \\ 
\emph{Vicia narbonensis} & 3 & 2\% & 1529 \\ 
\emph{Triticum urartu} & — & — & 2033 \\ 
\emph{Aegilops speltoides} & — & — & 1429 \\ 
\emph{Aegilops tauschii} & — & — & 1395 \\ 
\bottomrule

\begin{minipage}{\linewidth}
\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
\end{minipage}
\endgroup

Georeferenced occurrence data was obtained from the Global Biodiversity
Information Facility (GBIF) using via its application programming
interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF occurrences marked has
having imprecise or duplicate coordinates were excluded from the
training dataset, as were fossil records. Although niche models have
reasonable predictive power even with small training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 57 taxa with less than 50 occurrences in West Asia, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what account for such a disproportionately high number of
records.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia for the 56 modelled taxa (N=35208}

@@

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of \citet{BarbetMassinEtAl2012} for
regression-based species distribution models and use a large (:10000)
random sample of points, weighted equally against the presences in the
regression. \citet{ValaviEtAl2022} also recommend using a very large
background sample for random forest models.

@@

We modelled the occurrence of species as a function of X spatial
predictor variables (\textbf{?@tbl-predictors}). These included:

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(\textbf{?@tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for 56 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in \textbf{?@tbl-predictors}. Substituting
the ``current'' climate predictors for those derived from palaeoclimatic
simulations \citep{BrownEtAl2018}, we could then generate hindcast
predictions for reconstructed past environments in 4 key climate periods
-- a total of 224 modelled palaeodistributions. Predicted distributions
for individual taxa are presented in the appendix and accompanying
material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 98\%, with an average of 91\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 80\% and 100\%, with an average of 93\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.973±0.058 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Buglossoides
tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
\emph{Lathyrus sativus}, and \emph{Triticum turgidum dicoccum}. Test
metrics and ROC curves for the individual models are included in the
appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is significantly worse, with only 12\%
of presences in archaeobotanical assemblages successfully predicted.

\section{Results}\label{results}

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-range-density-1.pdf}}

@@

\caption{\label{fig-range-density}Distribution of predicted species
ranges by period. Dashed lines indicate the median range.}

@@

Species ranges are predicted to have been signicantly smaller in the
terminal Pleistocene and Early Holocene that under current conditions
Figure~\ref{fig-range-density}, though the magnitude of this change is
likely also to reflect a degree of overfitting in the model (see below).
Fluctuations in modelled range size between the Bølling-Allerød
(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
(11.7--8.3 ka) are more directly comparable. The average range of
modelled species shrank by 12\% over the Pleistocene--Holocene boundary,
with a noticeable dip in the Younger Dryas. This perhaps indicates that
although this period is considered one of climatatic amelioration
globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
may have supported more extensive plant-based economies in West Asia
specifically.

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
sativus} (both marginally edible legumes); \emph{Triticum aestivum
compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in
particular disappear from the predicted ranges of these species, leaving
the Levant and to a lesser extent the Aegean and Cyprus as refugia.

>>>>

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 81 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
The models predict significant reductions and/or shifts in species
ranges in the terminal Pleistocene and Early Holocene compared to
present conditions. Many species that are found throughout the region's
`hilly flanks' today are indicated to have much more restricted
distributions centered on the Levant, Cyprus and Western Anatolia. In
addition, overall ranges shrunk by an average of c.~25\% from the
terminal Pleistocene to the Early Holocene. Modelled ranges do not
reliably predict the observed occurrence of specific species at
archaeological sites, however {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    Introduction
  \item[$\square$]
    Background

@@

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-model-metrics
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\boxtimes$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

\caption{\label{fig-region}Late Epipalaeolithic \& Pre-Pottery Neolithic
archaeological sites used to generate modelled flora}

@@

The aim of our study was to model the biogeography of species relevant
to human subsistence economies in West Asia (excluding the Southern
Arabian peninsula, see Figure~\ref{fig-region}) during the
archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 96
taxa (Table~\ref{tbl-taxa}) comprising the identifiable species observed
at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites,
according to our previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. This was based on dataset of 117
assemblages collated from three previously published regional
archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
distinguish between the source of the remains; archaeobotanical
assemblages are subject to a variety of preservational and recovery
biases, so by no means were all the species on our list consumed or even
deliberately collected by people. However, we assume that there presence
at a site of human settlement at least implies that they were part of
the wider ecosystem that supported habitation there.

Taxonomic names were resolved to the canonical form specified in the
GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-taxa}). Domestic species meeting our inclusion criteria
were substituted with their wild progenitor(s), where different.

@@

Georeferenced occurrence data for West Asia was obtained from the Global
Biodiversity Information Facility (GBIF) using via its application
programming interface and the R package `rgbif'
\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
fossil occurrences, recorded absences, and records with missing,
imprecise (\textgreater1 km uncertainty), or invalid coordinates.
Although niche models have reasonable predictive power even with small
training samples
\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
excluded 81 taxa with less than 50 usable occurrences, following
recommendations for niche models generally and Random Forest-based
models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
as this would have been computationally prohibitive and we were
uncertain what accounted for such a disproportionately high number of
records.

@@

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia used to train models (N=78222}

@@

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of \citet{BarbetMassinEtAl2012} for
regression-based species distribution models and use a large (≈10000)
random sample of points, weighted equally against the presences in the
regression. \citet{ValaviEtAl2022} also recommend using a large
background sample for random forest models.

@@

We modelled the occurrence of species as a function of X spatial
predictor variables (Table~\ref{tbl-predictors}). These included:

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(Table~\ref{tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

@@

We trained Random Forest models for 81 taxa using contemporary
occurrence data from GBIF, a random sample of background points, and the
predictor variables described in Table~\ref{tbl-predictors}.
Substituting the ``current'' climate predictors for those derived from
palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
hindcast predictions for reconstructed past environments in 4 key
climate periods -- a total of 324 modelled palaeodistributions.
Predicted distributions for individual taxa are presented in the
appendix and accompanying material.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 99\%, with an average of 92\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 79\% and 100\%, with an average of 94\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.978±0.061 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Avena
fatua}, \emph{Lepidium perfoliatum}, \emph{Triticum turgidum dicoccum},
and \emph{Zygophyllum fabago}. Test metrics and ROC curves for the
individual models are included in the appendix.

The ability of the hindcast models to predict the occurrence of specific
species at archaeological sites is significantly worse, with only 5\% of
presences in archaeobotanical assemblages successfully predicted.

\begingroup
\setlength\LTleft{0\linewidth}
\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont

\begin{longtable}{@{\extracolsep{\fill}}lrrrrr}

\caption{\label{tbl-taxa}Summary of species modelled}

\toprule
 & \multicolumn{2}{c}{Occurrences} & \multicolumn{3}{c}{Model} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-6}
Taxon & Archaeological & Present (GBIF) & Accuracy & ROC-AUC & Sensitivity \\ 
\midrule\addlinespace[2.5pt]
{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.99 \\ 
{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.89 \\ 
{\itshape Aegilops speltoides} & — & 701 & 0.93 & 0.98 & 0.94 \\ 
{\itshape Aegilops tauschii} & — & 725 & 0.90 & 0.98 & 0.94 \\ 
{\itshape Aizoanthemopsis hispanica} & 4 & 515 & 0.98 & 0.99 & 0.96 \\ 
{\itshape Ammi majus} & 4 & 900 & 0.99 & 0.99 & 0.98 \\ 
{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.96 & 0.88 \\ 
{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.85 & 0.94 & 0.91 \\ 
{\itshape Arnebia decumbens} & 21 & 324 & 0.94 & 0.98 & 0.93 \\ 
{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Asphodelus aestivus} & 2 & 76 & 0.96 & 0.97 & 0.96 \\ 
{\itshape Atriplex prostrata} & 4 & 266 & 0.97 & 1.00 & 0.97 \\ 
{\itshape Avena fatua} & 2 & 579 & 0.85 & 0.93 & 0.85 \\ 
{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Bolboschoenus maritimus} & 29 & 1040 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 \\ 
{\itshape Bromus sterilis} & 3 & 759 & 0.96 & 1.00 & 0.98 \\ 
{\itshape Buglossoides arvensis} & 23 & 260 & 0.86 & 0.96 & 0.90 \\ 
{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.94 \\ 
{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 \\ 
{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.93 \\ 
{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.92 \\ 
{\itshape Cicer reticulatum} & 3 & 52 & 0.96 & 1.00 & 0.92 \\ 
{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.95 & 0.89 \\ 
{\itshape Crithopsis delileana} & 3 & 406 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Erodium ciconium} & 2 & 359 & 0.91 & 0.99 & 0.99 \\ 
{\itshape Ficus carica} & 8 & 3495 & 0.93 & 0.98 & 0.93 \\ 
{\itshape Fumaria densiflora} & 4 & 413 & 0.98 & 0.99 & 0.98 \\ 
{\itshape Gypsophila elegans} & 3 & 76 & 0.93 & 0.96 & 0.85 \\ 
{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.99 & 0.95 \\ 
{\itshape Gypsophila vaccaria} & 8 & 511 & 0.96 & 0.99 & 0.93 \\ 
{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.99 & 1.00 & 1.00 \\ 
{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.98 & 0.96 \\ 
{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.98 & 0.99 & 0.97 \\ 
{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 \\ 
{\itshape Hordeum murinum} & 5 & 1949 & 0.95 & 0.99 & 0.96 \\ 
{\itshape Hordeum spontaneum} & 51 & 4656 & 0.96 & 0.99 & 0.97 \\ 
{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.94 \\ 
{\itshape Lathyrus aphaca} & 4 & 2079 & 0.95 & 0.99 & 0.96 \\ 
{\itshape Lathyrus cicera} & 2 & 591 & 0.90 & 0.98 & 0.94 \\ 
{\itshape Lathyrus oleraceus} & 6 & 1580 & 0.88 & 0.97 & 0.94 \\ 
{\itshape Lathyrus sativus} & 4 & 196 & 0.80 & 0.94 & 0.87 \\ 
{\itshape Lepidium perfoliatum} & 3 & 83 & 0.80 & 0.92 & 0.83 \\ 
{\itshape Linum bienne} & 9 & 376 & 0.96 & 1.00 & 0.98 \\ 
{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Lolium temulentum} & 3 & 159 & 0.93 & 0.95 & 0.92 \\ 
{\itshape Medicago astroites} & 15 & 143 & 0.90 & 0.97 & 0.88 \\ 
{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.98 \\ 
{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.97 \\ 
{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.99 & 1.00 \\ 
{\itshape Neotorularia torulosa} & 2 & 539 & 0.98 & 0.99 & 0.96 \\ 
{\itshape Peganum harmala} & 2 & 917 & 0.91 & 0.98 & 0.95 \\ 
{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 0.99 & 0.98 \\ 
{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.97 \\ 
{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.98 \\ 
{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.90 \\ 
{\itshape Prosopis farcta} & 5 & 2263 & 0.97 & 1.00 & 0.97 \\ 
{\itshape Quercus ithaburensis} & 2 & 2216 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Rumex pulcher} & 6 & 1394 & 0.98 & 1.00 & 0.99 \\ 
{\itshape Salsola kali} & 6 & 537 & 0.98 & 1.00 & 1.00 \\ 
{\itshape Salvia absconditiflora} & 3 & 105 & 0.92 & 0.99 & 0.91 \\ 
{\itshape Secale cereale} & 4 & 268 & 0.79 & 0.95 & 0.92 \\ 
{\itshape Secale strictum} & 3 & 124 & 0.74 & 0.96 & 0.92 \\ 
{\itshape Stipa dregeana} & 2 & 79 & 0.93 & 0.99 & 0.97 \\ 
{\itshape Suaeda fruticosa} & 3 & 238 & 0.99 & 1.00 & 1.00 \\ 
{\itshape Taeniatherum caput-medusae} & 4 & 262 & 0.82 & 0.96 & 0.92 \\ 
{\itshape Tribulus terrestris} & 2 & 1091 & 0.91 & 0.97 & 0.93 \\ 
{\itshape Triticum aestivum compactum} & 2 & 147 & 0.80 & 0.96 & 0.92 \\ 
{\itshape Triticum durum} & 3 & 95 & 0.90 & 0.94 & 0.88 \\ 
{\itshape Triticum monococcum aegilopoides} & 36 & 1176 & 0.88 & 0.98 & 0.96 \\ 
{\itshape Triticum turgidum dicoccum} & 47 & 60 & 0.72 & 0.85 & 0.79 \\ 
{\itshape Triticum urartu} & — & 424 & 0.93 & 1.00 & 1.00 \\ 
{\itshape Verbena officinalis} & 3 & 934 & 0.95 & 0.99 & 0.96 \\ 
{\itshape Vicia ervilia} & 26 & 688 & 0.88 & 0.97 & 0.93 \\ 
{\itshape Vicia faba} & 7 & 1597 & 0.91 & 0.98 & 0.95 \\ 
{\itshape Vicia narbonensis} & 3 & 1451 & 0.93 & 0.98 & 0.94 \\ 
{\itshape Vicia orientalis} & 14 & 353 & 0.84 & 0.95 & 0.88 \\ 
{\itshape Zygophyllum fabago} & 3 & 263 & 0.89 & 0.95 & 0.84 \\ 
\bottomrule

\endgroup

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pred-range-density-1.pdf}}

@@

\caption{\label{fig-pred-range-density}Distribution of predicted species
ranges by period. Dashed lines indicate the median range.}

@@

Our reconstructed palaeodistributions (shown in full in the appendix)
indicate that the majority of species have experienced a significant
reduction and/or shift in their ranges since the Early Holocene. 75 of
81 species have reduced ranges; 65 of more than 10\% or more. Though the
magnitude of the reduction from prehistory to the present likely
reflects a degree of overfitting in the model (discussed further below),
fluctuations in modelled range size between the Bølling-Allerød
(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
(11.7--8.3 ka) are more directly comparable
\textbf{?@fig-range-density}. The average range of modelled species was
25\% in the Early Holocene compared to the Bølling--Allerød, and 32\%
during the Younger Dryas (i.e.~ranges recovered slightly between the
Younger Dryas and Early Holocene). This perhaps indicates that although
this period is considered one of climatatic amelioration globally
\citep{JonesEtAl2019}, the colder conditions of the Pleistocene may have
supported more extensive plant-based economies in West Asia
specifically.

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-sum-1.pdf}}

\caption{\label{fig-palaeodist-sum}Predicted species richness (sum of
predicted ranges) by period}
